Coupled response of stomatal and mesophyll conductance to light enhances photosynthesis of shade leaves under sunflecks

### 2016-09-22 PCE_paper/review_response.Rmd

\
*Reviewer: 8. line 404 and 405: compare to Brooks’s study mentioned above and expand to address fixed vs. variable controls*

\
*Response*: - line 404-405: We added the Brooks et al. 1996 citation explaining observed patterns of morphology and physiology when sun leaves become shaded during tree growth and crown development.

\
*Reviewer: 9. line 412 to 415: if I recall correctly, optimization of WUE in the presence of similar hydraulics should lead to similar WUE throughout the canopy. Is this true? If so, it should be stated and the discrepancy made clearer. The point is made (I think), but it should be crystal clear that the optimization arguments are failing us here, which may mean that we’re missing something. Approximately this point is made on lines 431 to 434, but as presented it seems like a minor point. I think it is not so minor.*

\
*Response*: If the cost of water (sensu Cowan & Farquhar 1977) is constant throughout the canopy, then we would expect similar WUE in sun and shade leaves. Since we found that shade leaves have lower WUE, this could indicate a lower cost of water. This is consistent with the observation that hydraulic conductance was higher in shade leaves. We have expanded the section (previously lines 431-434) to read:

"For the entire canopy to exhibit optimal regulation of stomatal conductance, that is, to maximize photosynthesis minus transpiration cost (Cowan and Farquhar 1977), sun and shade leaves should have equal ITE - provided environmental drivers are similar (Buckley et al. 2014). In our study, higher *g*~s~ in shade leaves may in part be explained by higher *K*~L~, which effectively reduces the cost of water transport, allowing more profligate water use in shade leaves. Consistent with our results, Buckley et al. (2014) found that shade leaves in a grapevine canopy exhibited higher *g*~s~ than expected based on their environment and comparison to sun leaves. Further work is needed to confirm the generality of this observation."

### 2016-09-22 PCE_paper/manuscript_gmes.Rmd
colors finished

##Whole-tree chamber experimental design
Twelve *Eucalyptus tereticornis* Sm. seedlings, chosen from a single local Cumberland plain cohort, were planted in March 2013 into 12 whole-tree chambers (WTC) at the Hawkesbury Forest Experiment site near Richmond, New South Wales, Australia. The top soils at this site, used in the chambers, are an alluvial formation of low-fertility sandy loam soils (380 and 108 mg kg^-1^ total N and phosphorus respectively) with low organic matter (0.7 %) and low water holding capacity. A root exclusion barrier extended from chamber walls to the hard layer (ca. 1 m) and roots were allowed to grow freely below the barrier. Each chamber was 9 m tall, which accommodated growth of trees for 15 months. A detailed description of the WTC operation and design is available in Barton et al. [-@barton2010whole] and methods for this experiment in Drake et al. [-@drake2016does]. Coppiced *E. saligna* trees were present in the area between the whole-tree chambers, and these coppiced trees grew at rates similar to the trees in the chambers, thus providing ample shade. Six chambers were set to match outside ambient air temperatures (AT) while the remaining 6 experienced an elevated air temperature treatment of ambient +3&deg;C (ET, Figure `r airvarfig`). The trees grew quickly and developed large canopies, with height growth reaching the top of the WTCs at the end of the experiment. Trees were watered weekly with 70 L from March 2013 to September 2013. From October 2013 to the end of the experiment trees were watered every 15 days with the mean monthly rainfall amount for Richmond, NSW. In February 2014 half of the chambers (3 each of AT and ET) were subjected to a drought treatment by withholding watering. Because of the difficulty of measuring *g*~m~ in the droughted trees (the necessary CO~2~ drawdown was not achieved), we here report only data from well-watered trees, which reduces the sample size from *n*=6 to *n*=3 for the final 3 months of the experiment.

@@

\
Concurrent gas exchange and C isotope discrimination measurements were conducted based on methods described in Tazoe et al. [-@tazoe2011using] and Evans & von Caemmerer [-@evans2013temperature]. Leaf level gas exchange was measured with a standard (2 x 3 cm) leaf cuvette and red/blue light source using a portable gas exchange system (LI-6400XT, LI-COR, Lincoln, NE, USA). This system was coupled with a tunable diode laser (TDL; TGA100,Campbell Scientific, Inc., Logan, UT, USA) for concurrent measurements of ^13^C isotope discrimination. The CO~2~ in the leaf cuvette was set at ambient atmospheric [CO~2~] (400 ppm) with a flow rate of 200 $\mu$mol s^-1^. Two identical gas exchange systems were run simultaneously, one in each of a randomly chosen WTC for each temperature treatment. Leaf temperatures were controlled at the current AT or ET WTC air temperature, and measurements were made at ambient air humidity. Measurements were thus made across a range of air temperatures and vapour pressure deficits (kPa) across the six campaigns. PPFD in the cuvette was set to match the individual light environment of each leaf type (explained above). Sustained periods of high irradiance (sunflecks) were simulated for shade leaves by increasing the leaf cuvette PPFD (LI-COR red/blue light source) to match the light environment of the sun leaf in the same tree. The maximum sunfleck response of shade leaves was then recorded once CO~2~ and water vapour fluxes re-stabilized in the leaf cuvette (ca. 25 min).

@@

\
Mesophyll conductance was calculated from carbon isotope discrimination with equations and fractionation factors as presented in Evans & von Caemmerer [-@evans2013temperature], including the ternary corrections proposed by Farquhar & Cernusak [-@farquhar2012ternary], such that:

@@

where $\Delta$~o~ is the observed discrimination and $\Delta$~i~, $\Delta$~gm~ , $\Delta$~e~ and $\Delta$~f~ are the contributions to fractionation if *C*~i~ = *C*~c~, *g*~m~, respiration and photorespiration, respectively. For this study, the CO~2~ compensation point ($\Gamma_{*}$) and respiration during the day (R~d~) parameters originally derived for tobacco plants [@von1994kinetics] were replaced with parameters derived for *Eucalyptus globulus* from Crous *et al*. [-@crous2012light] when calculating *g*~m~. Full descriptions of the carbon isotope discrimination equations, with ternary corrections, are presented in Supporting Information Methods S1. The variation in $\Delta$~o~ between sun and shade leaves and the simulated sunfleck were compared as a function of *C*~i~/C~a~. Once *g*~m~ was calculated, the chloroplast CO~2~ partial pressure (*C*~c~) and the drawdown of CO~2~ from the intercellular air spaces to the site of carboxylation were estimated from the relationship between *g*~m~ and leaf photosynthesis rate (*A*~n~) by:

@@

##Data analysis
Differences in responses of dependent variables to either the warming treatment or leaf type were analysed with linear mixed-effects models with WTC as a random effect, and leaf type (sun, shade or shade in the sun) and warming treatment, and the interaction as fixed effects. The interaction was never significant, so we only report results of the tests of the main effects. Explained variance (*R*^2^) of mixed models was computed as in Nakagawa & Schielzeth [-@nakagawa2013general], in which the marginal *R*^2^ represents variance explained by fixed factors and the conditional *R*^2^ by both fixed and random factors. Confidence intervals (95 %) of linear mixed-effects models were generated using bootstrapping methods with 999 simulations, using the 'bootMer' function in the 'lme4' package [@bates2013lme4]. For non-linear relationships, confidence intervals were estimated by fitting a generalized additive model to the data with the 'mgcv' package [@wood2006], using WTC as a random effect. All tests of statistical significance were conducted at an $\alpha$ of 0.05. For multiple post-hoc comparisons, we used Tukey pairwise comparison tests with the 'multcomp' package [@hothorn2008]. In all results, we report the mean &pm; one standard error for the mean (SE). All analyses were performed with R 3.3.0 [@RDevelopmentCoreTeam2011].

@@

\
Photosynthesis is also limited by the supply of water to the upper canopy. Ultimately, the ability of tree hydraulic architecture to supply water to foliage across increasing pathlengths affects productivity and survival [@sellin2008effects]. Using a two-leaf model, Peltoniemi et al. [-@peltoniemi2012co] theorized that optimal N distribution will be proportional to light distribution only if *K*~L~ is also optimally distributed. We found higher *K*~L~ in shade leaves compared to sun leaves, contrary to the prediction by Peltoniemi et al. [-@peltoniemi2012co], but consistent with higher *g*~s~ attained by shade leaves in high light conditions. This finding is consistent with the idea that *g*~s~ is regulated, at least in part, by *K*~L~ [@hubbard2001stomatal].

@@

\
In sun leaves, *A*~n~ and *g*~s~ were strongly correlated, exhibiting behaviour broadly consistent with optimal stomatal theory (Fig. 5). However, lower rates of *A*~n~ in shade leaves were not coupled with decreases in *g*~s~, leading to the observed decreases in ITE. This is significant as optimal stomatal regulation to balance C gain with water loss has been reported across a wide range of ecosystems and plant functional types; however, empirical data is often collected only on sun leaves [e.g. @prentice2014balancing; @lin2015optimal]. For the entire canopy to exhibit optimal regulation of stomatal conductance, that is, to maximize photosynthesis minus transpiration cost [@cowan1977stomatal], sun and shade leaves should have equal ITE - provided environmental drivers are similar [@buckley2014]. In our study, higher *g*~s~ in shade leaves may in part be explained by higher *K*~L~, which effectively reduces the cost of water transport, allowing more profligate water use in shade leaves. Consistent with our results, Buckley et al. [-@buckley2014] found that shade leaves in a grapevine canopy exhibited higher *g*~s~ than expected based on their environment and comparison to sun leaves. Further work is needed to confirm the generality of this observation.

@@

**Figure 2**. Photosynthetic capacity (*A*~max~) was positively correlated with mean leaf N per area (*N*~a~) for sun and shade leaves grown under elevated (ET) and ambient (AT) temperature treatments. Leaf *N*~a~ is the mean of each leaf type sampled across gas exchange campaigns in each whole-tree chamber. *A*~max~ was measured once on all trees for each leaf type, in February 2014, at 25°C and at both saturating light and [CO~2~]. The dashed line represents the significant linear model fit for all leaves, with a marginal and conditional *R*^2^ of 0.59 and 0.62.

>>>>

##Whole-tree chamber experimental design
Twelve *Eucalyptus tereticornis* Sm. seedlings, chosen from a single local Cumberland plain cohort, were planted in March 2013 into 12 whole-tree chambers (WTC) at the Hawkesbury Forest Experiment site near Richmond, New South Wales, Australia. The top soils at this site, used in the chambers, are an alluvial formation of low-fertility sandy loam soils (380 and 108 mg kg^-1^ total N and phosphorus respectively) with low organic matter (0.7 %) and low water holding capacity. A root exclusion barrier extended from chamber walls to the hard layer (ca. 1 m) and roots were allowed to grow freely below the barrier. Each chamber was 9 m tall, which accommodated growth of trees for 15 months. A detailed description of the WTC operation and design is available in @barton2010whole and methods for this experiment in @drake2016does. Coppiced *E. saligna* trees were present in the area between the whole-tree chambers, and these coppiced trees grew at rates similar to the trees in the chambers, thus providing ample shade. Six chambers were set to match outside ambient air temperatures (AT) while the remaining 6 experienced an elevated air temperature treatment of ambient +3&deg;C (ET, Figure `r airvarfig`). The trees grew quickly and developed large canopies, with height growth reaching the top of the WTCs at the end of the experiment. Trees were watered weekly with 70 L from March 2013 to September 2013. From October 2013 to the end of the experiment trees were watered every 15 days with the mean monthly rainfall amount for Richmond, NSW. In February 2014 half of the chambers (3 each of AT and ET) were subjected to a drought treatment by withholding watering. Because of the difficulty of measuring *g*~m~ in the droughted trees (the necessary CO~2~ drawdown was not achieved), we here report only data from well-watered trees, which reduces the sample size from *n*=6 to *n*=3 for the final 3 months of the experiment.

@@

\
Concurrent gas exchange and C isotope discrimination measurements were conducted based on methods described in @tazoe2011using and @evans2013temperature. Leaf level gas exchange was measured with a standard (2 x 3 cm) leaf cuvette and red/blue light source using a portable gas exchange system (LI-6400XT, LI-COR, Lincoln, NE, USA). This system was coupled with a tunable diode laser (TDL; TGA100,Campbell Scientific, Inc., Logan, UT, USA) for concurrent measurements of ^13^C isotope discrimination. The CO~2~ in the leaf cuvette was set at ambient atmospheric [CO~2~] (400 ppm) with a flow rate of 200 $\mu$mol s^-1^. Two identical gas exchange systems were run simultaneously, one in each of a randomly chosen WTC for each temperature treatment. Leaf temperatures were controlled at the current AT or ET WTC air temperature, and measurements were made at ambient air humidity. Measurements were thus made across a range of air temperatures and vapour pressure deficits (kPa) across the six campaigns. PPFD in the cuvette was set to match the individual light environment of each leaf type (explained above). Sustained periods of high irradiance (sunflecks) were simulated for shade leaves by increasing the leaf cuvette PPFD (LI-COR red/blue light source) to match the light environment of the sun leaf in the same tree. The maximum sunfleck response of shade leaves was then recorded once CO~2~ and water vapour fluxes re-stabilized in the leaf cuvette (ca. 25 min).

@@

\
Mesophyll conductance was calculated from carbon isotope discrimination with equations and fractionation factors as presented in @evans2013temperature, including the ternary corrections proposed by @farquhar2012ternary, such that:

@@

where $\Delta$~o~ is the observed discrimination and $\Delta$~i~, $\Delta$~gm~ , $\Delta$~e~ and $\Delta$~f~ are the contributions to fractionation if *C*~i~ = *C*~c~, *g*~m~, respiration and photorespiration, respectively. *C*~i~ and  *C*~c~ are the intercellular and chloroplast CO~2~ partial pressures. For this study, the CO~2~ compensation point ($\Gamma_{*}$) and respiration during the day (R~d~) parameters originally derived for tobacco plants [@von1994kinetics] were replaced with parameters derived for *Eucalyptus globulus* from @crous2012light when calculating *g*~m~. Full descriptions of the carbon isotope discrimination equations, with ternary corrections, are presented in Supporting Information Methods S1. The variation in $\Delta$~o~ between sun and shade leaves and the simulated sunfleck were compared as a function of *C*~i~/C~a~. Once *g*~m~ was calculated, *C*~c~ and the drawdown of CO~2~ from the intercellular air spaces to the site of carboxylation were estimated from the relationship between *g*~m~ and leaf photosynthesis rate (*A*~n~) by:

@@

##Data analysis
Differences in responses of dependent variables to either the warming treatment or leaf type were analysed with linear mixed-effects models with WTC as a random effect, and leaf type (sun, shade or shade in the sun) and warming treatment, and the interaction as fixed effects. The interaction was never significant, so we only report results of the tests of the main effects. Explained variance (*R*^2^) of mixed models was computed as in @nakagawa2013general, in which the marginal *R*^2^ represents variance explained by fixed factors and the conditional *R*^2^ by both fixed and random factors. Confidence intervals (95 %) of linear mixed-effects models were generated using bootstrapping methods with 999 simulations, using the 'bootMer' function in the 'lme4' package [@bates2013lme4]. For non-linear relationships, confidence intervals were estimated by fitting a generalized additive model to the data with the 'mgcv' package [@wood2006], using WTC as a random effect. All tests of statistical significance were conducted at an $\alpha$ of 0.05. For multiple post-hoc comparisons, we used Tukey pairwise comparison tests with the 'multcomp' package [@hothorn2008]. In all results, we report the mean &pm; one standard error for the mean (SE). All analyses were performed with R 3.3.0 [@RDevelopmentCoreTeam2011].

@@

\
Photosynthesis is also limited by the supply of water to the upper canopy. Ultimately, the ability of tree hydraulic architecture to supply water to foliage across increasing pathlengths affects productivity and survival [@sellin2008effects]. Using a two-leaf model, @peltoniemi2012co theorized that optimal N distribution will be proportional to light distribution only if *K*~L~ is also optimally distributed. We found higher *K*~L~ in shade leaves compared to sun leaves, contrary to the prediction by @peltoniemi2012co, but consistent with higher *g*~s~ attained by shade leaves in high light conditions. This finding is consistent with the idea that *g*~s~ is regulated, at least in part, by *K*~L~ [@hubbard2001stomatal].

@@

\
In sun leaves, *A*~n~ and *g*~s~ were strongly correlated, exhibiting behaviour broadly consistent with optimal stomatal theory (Fig. 5). However, lower rates of *A*~n~ in shade leaves were not coupled with decreases in *g*~s~, leading to the observed decreases in ITE. This is significant as optimal stomatal regulation to balance C gain with water loss has been reported across a wide range of ecosystems and plant functional types; however, empirical data is often collected only on sun leaves [e.g. @prentice2014balancing; @lin2015optimal]. For the entire canopy to exhibit optimal regulation of stomatal conductance, that is, to maximize photosynthesis minus transpiration cost [@cowan1977stomatal], sun and shade leaves should have equal ITE - provided environmental drivers are similar [@buckley2014]. In our study, higher *g*~s~ in shade leaves may in part be explained by higher *K*~L~, which effectively reduces the cost of water transport, allowing more profligate water use in shade leaves. Consistent with our results, @buckley2014 found that shade leaves in a grapevine canopy exhibited higher *g*~s~ than expected based on their environment and comparison to sun leaves. Further work is needed to confirm the generality of this observation.

@@

**Figure 2**. Photosynthetic capacity (*A*~max~) was positively correlated with mean leaf N per area (*N*~a~) for sun and shade leaves grown under elevated (ET) and ambient (AT) temperature treatments. Leaf *N*~a~ is the mean of each leaf type sampled across six gas exchange campaigns in each whole-tree chamber. *A*~max~ was measured once on all trees for each leaf type, in February 2014, at 25°C and at both saturating light and [CO~2~]. The dashed line represents the significant linear model fit for all leaves, with a marginal and conditional *R*^2^ of 0.59 and 0.62.
